Domain-specific working memory loads selectively increase negative interpertations of surprised facial expressions

### 2019-10-01 Manuscript/Manuscript.Rmd
reorganized intro

---
title             : "Domain-specific working memory loads selectively increase negative interpertations of surprised facial expressions"
shorttitle        : "DOMAIN-SPECIFIC WORKING MEMORY AND SURPRISED EXPRESSIONS"

@@

### load necessary libraries ###
suppressPackageStartupMessages(library(readxl)) 
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(openxlsx))
suppressPackageStartupMessages(library(Rmisc))
suppressPackageStartupMessages(library(broom))
```
# Introduction
## Working memory and load theory
Despite extensive research on the interaction of working memory and affective processes, there is much to learn concerning how cognitive and emotional processes affect one another. Executive functions, including working memory, are related to successful self-regulation, and in turn emotion regulation [@hofmann_executive_2012]. Directly comparing working memory and self-regulation of emotional responses, Schmeichel and colleagues [-@schmeichel_working_2008] reported that individuals with higher levels of working memory capacity demonstrated improved self-regulation towards the emotional stimuli. This suggests a connection--perhaps through some shared resource pool--between mitigated emotional responding and larger working memory resource availablility. Other work has focused on the effects of moods or affective states on working memory performance. For instance, some reports claim that both positive and negative mood interfere with working memory (Eyesenck and Calvo, 1992); however, others suggest benefits of positive mood on working memory [@yang_positive_2013]. Similarly, active working memory processes may alter concurrent affective processes. For instance, actively engaging working memory can mitigate emotional responses, particularly to negative stimuli. Recent neuroimaging work reports that negative emotional responses decrease as the cognitive demands of a working memory task increase [@van_dillen_tuning_2009]. Additionally, following an anger induction, those with low trait rumination show faster blood pressure recovery when provided with a distractor task [@gerin_role_2006]. Together, these studies suggest a resource competition between cognitive and emotional processes; in other words, when cognitive load demands are high (e.g., during active working memory maintenance), there are fewer resources available for other (i.e., affective) processes.

While previous work suggests an inhibitory effect on some emotional responses during periods with high cognitive demands, researchers have primarily focused on emotional responses to clearly valenced emotional stimuli. For instance, Schmeichel and colleagues [-@schmeichel_working_2008] showed participants videos intended to elicit strong negative responses (e.g., disgust) or positive responses (e.g., humor), while others have focused on comparing responses to neutral compared to negative sitmuli [@van_dillen_tuning_2009].<!--  Others have even measured symptoms during working memory maintenance of affective symptoms (i.e., typically increased negative affect), such as anxiety [@spangler_little_2017].--> However, many emotional appraisals in day-to-day life are more nuanced than those invoked by many of the images of negative stimuli one might encounter in the lab (e.g., snakes, mutilated bodies). For example, one may appraise the content of a billboard displaying a large order of french fries as either negative or positive depending on whether or not consuming that food is (in)congruent with one's current goals. This emotional appraisal is completed under concurrent load demands--that is, the perceiver must process both the emotional stimulus (i.e., the fries) as well as actively maintain their current goal state. Maintenance of one's goal state, such as a diet, can reduce performance on some components of executive functioning tasks [@shaw_dieting_2004]. These observations are in line with Lavie and colleague's [-@lavie_load_2004] load theory, which posits that under a large cognitive load less executive resources are available to regulate incoming stimulus information. <!-- In turn, this could mean that more readily processed perceptual input (e.g., low spatial frequency images; LSF) might drive decisions under working memory loads. -->

Recently, cognitive load theory researchers have tested, and demonstrated, the domain-specificity of load and distractor interference in visual, spatial, and phonological domains [@burnham_components_2014]. Load interference effects may also transverse other domain componenets, such as emotional compared to neutral memory content. Emotional stimuli readily capture attention compared to neutral stimuli, and this is true even in participants with amygdala damage [@hodsoll_attentional_2011; @piech_attentional_2011]. Given emotional stimuli's priority position in the information processing stream, it may be that cognitive loads with emotional content, compared to neutral, differentially affect concurrent emotional appraisals. Indeed, Kensigner and colleagues [-@kensinger_effect_2003] showed that negative emotional content slows performance on the n-back task. Other domain-specific effects have been observed in many lines of executive functions research, including those beyond the working memory domain. For example, the Stroop task [@stroop_studies_1935], a common measurement tool for inhibitory control, has been modified to include both emotional and non-emotional (neutral) stimuli [@whalen_emotional_2006] which has pronounced effects when the emotional words are population specific (e.g., trauma words in a PTSD sample). Other neuroimaging work also supports the notion that separate systems handle attentional biasing for domain-specific (emotional vs. non-emotional) task relevancy [@egner_dissociable_2008]. Given this evidence for dissociations of emotional and non-emotional information domains in executive functions (e.g., working memory, inhibitory control), the present work aims to clarify the interaction of emotional and non-emotional visual working memory demands on concurrent interpretations of emotional stimuli with ambiguous valence.

@@

## Interpreting ambiguity
<!-- Facial expressions are rich with affective information, and correctly interpreting these social cues is critical for successfully navigating the social world.--> Individuals differ in their tendency to interpret ambiguously valenced stimuli, like a tempting food item or a surprised facial expression, as either positive or negative. <!-- Additionally, facial expressions can serve as ambiguously valenced stimuli. While most facial expressions send a clear social signal, this is not always the case. While a smile likely expresses a positive affective state, other cues are not so clear.--> This is attributable to these stimuli's predictive value for both positive and negative outcomes. For instance, a surprised expression could signal positive (e.g., winning the lottery) or negative (e.g., a car accident) events. This affective bias is known as one's *valence bias*, and a growing body of work has used both facial expressions and scenes to quantify this individual difference [@neta_corrugator_2009; @neta_primacy_2010; @neta_neural_2013]. Chronic negativity biases in memory and attention are related to psychopathology, such as depression and anxiety [@mathews_cognitive_2005], suggesting the importance of understanding the factors that contribute to individuals' biases. Importantly, the valence bias is a stable measure with participants showing positively correlated scores across a one year time gap [@neta_corrugator_2009]. Despite the relative stability of the measure, experimental manipulations are capable of shifting an individuals bias [@brown_cortisol_2017; @neta_through_2018; @neta_its_2018].

Myriad factors contribute to an individual's bias, but the initial interpretation is thought to be negative across individuals. Data supporting this initial negativity hypothesis come from many studies. For instance, reaction times are faster for negative interpretations of ambiguous stimuli [@neta_dont_2016]. Additionally, presentation of surprised facial expressions as low spatial frequency images, which are processed more readily than high spatial frequency images, biased interpretations towards negativity [@neta_primacy_2010]. Under this framework, arriving at a positive interpretation requires additional, top-down regulatory processes, and there is evidence to support this as well. For example, forcing participants to slow their responding during interpretations of ambiguous images shifts individuals' biases towards positivity [@neta_its_2018]. Evidence from the neuroimaging literature supports the initial negativity hypothesis as well; more positive individuals show higher levels of BOLD activation in brain regions recruited during emotion regulation [@petro_individual_2018]. Perceptual input also contributes to valence bias. In one recent study, Neta and colleagues [-@neta_all_2017] showed that faster intial fixation on the mouth is related to more positive interpretations of surprised faces [@neta_all_2017].  Further, forcing gaze patterns to match those of the participants with the most negative or positive bias modulated interpretations of surprised expressions [@neta_through_2018]. In all, valence bias is a useful metric for understanding both trait-like components of individuals' affective biases, as well as gauging the effects of other experimental maniuplations on affective biases. 
<!--For instance, a surprised expression could signal either a positive (e.g., winning the lottery) or negative (e.g., seeing a snake in the woods) affective state in the expresser. When contextual information is limited, individuals differ in their tendency to interpret surprised facial expressions as positive or negative. Importantly, this affective bias extends beyond facial expressions, as individuals often show a similar bias to surprised faces as they do for ambiguous scenes [@neta_neural_2013] or even words (Harp, Petro, Brown & Neta, in prep). This bias towards positive or negative interpretations is known as one's valence bias and myriad factors contribute to one's bias. -->
<!-- Both bottom-up (e.g., perceptual input) and top-down (e.g., stereotypes) processes are recruited for interpretation of facial expressions, and a growing body of work suggests that the initial interpretation of emotionally ambiguous stimuli is negative and driven by bottom-up processes. For instance, reaction times are faster for negative interpretations of ambiguous stimuli (just faces??) [@neta_dont_2016]. Additionally, presentation of surprised facial expressions as low spatial frequency images, which is processed more readily than high spatial frequency images, biased interpretations towards negativity [@neta_primacy_2010]. Consequently, under this framework, arriving at a positive interpretation requires additional, top-down regulatory processes and there is evidence to support this as well. For example, forcing participants to slow their responding during interpretations of ambiguous images shifts individuals' biases towards positivity [@neta_its_2018]. Neuroimaging evidence supports this initial negativity hypothesis as well, more positive individuals show higher levels of BOLD activation in brain regions recruited during emotion regulation [@petro_individual_2018]. <!-- Perceptual input also contributes to valence bias. In one recent study, Neta and colleagues [-@neta_all_2017] showed that faster intial fixation, as well as longer overall fixation, on the mouth is related to more positive interpretations of surprised faces and that forcing gaze patterns to match those of modulated interpretations of surprised expressions. THIS NEEDS TO BE REWORKED In short, slowing response time allows individuals to better regulate the potentially negative information in surprised expressions and to see it in a more positive light.  -->

Given the evidence that a regulatory mechanism is necessary for positive interpretations of ambiguity, a demanding cognitive load might interfere with successful regulation. Indeed, Mattek and colleagues [-@mattek_differential_2016] recently showed that high levels of cognitive load (i.e., holding either a single or seven digit number in working memory) mitigates mouse trajectory deviations to the modal response. However, there was no effect on subjective valence interpretations. Just as phonological and visual cognitive loads and distractors showed domain-specificity [@burnham_components_2014], and that domain-speicificity is observed in BOLD data [@egner_dissociable_2008], a domain-specific cognitive load may differentially affect interpretations of ambiguity. In other words, there is likely a relationship between the qualities of stimuli held in working memory and effects on concurrent task processing. Here, we aim to test the effects of low and high working memory loads in both emotional and neutral domains on valence bias. We expect that trials in which participants are maintaining an emotional working memory load will be more negative than neutral trials, as the high cognitive demand interferes with the regulatory mechanisms used to arrive at a positive interpretation. Further, we predict that higher working memory laod trials, specifically in the emotional domain, will result in even more exaggerated negative interpretations.

>>>>

---
title             : "Domain-specific working memory loads selectively increase negative interpretations of surprised facial expressions"
shorttitle        : "DOMAIN-SPECIFIC WORKING MEMORY AND SURPRISED EXPRESSIONS"

@@

### load necessary libraries ###
suppressPackageStartupMessages(library(readxl)) 
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(openxlsx))
suppressPackageStartupMessages(library(Rmisc))
suppressPackageStartupMessages(library(broom))
```
# Introduction
## Facial expressions and individual differences
<!-- Facial expressions are rich with affective information, and correctly interpreting these social cues is critical for successfully navigating the social world.--> Facial expressions are a mechanism for communicating emotion during social interactions. While the use of facial expressions as an information exchange may be universal [@izard_innate_1994], there are notable differences in interpretations of facial expressions [@green_factors_2018; @neta_corrugator_2009]. One such difference is in valence judgments of emotional facial expressions. The affective circumplex model posits that two independent neurophysiological systems give rise to emotional experience which are known as valence and arousal [@posner_circumplex_2005]. Valence is the inherent positive (i.e., pleasant) or negative (i.e., unpleasant) qualities of a stimulus, feeling, or state. Importantly, individuals differ in their tendency to interpret ambiguously valenced stimuli, such as a tempting food item or a surprised facial expression, as either positive or negative. <!-- Additionally, facial expressions can serve as ambiguously valenced stimuli. While most facial expressions send a clear social signal, this is not always the case. While a smile likely expresses a positive affective state, other cues are not so clear.--> This is attributable to such stimuli's predictive value for both positive and negative outcomes in an individual's previous experience. For instance, a surprised expression could signal positive (e.g., winning the lottery) or negative (e.g., a car accident) events. This affective bias is known as one's *valence bias*, and a growing body of work has used both facial expressions and scenes to better understand this individual difference [@neta_corrugator_2009; @neta_primacy_2010; @neta_neural_2013].

Previous work shows that valence bias is both a stable, trait-like measure [@neta_corrugator_2009], but also malleable due to its susceptibility to laboratory manipulations [@brown_cortisol_2017; @neta_through_2018; @neta_its_2018]. The trait-like nature of valence bias is useful for understanding chronic affective biases. Negativity biases, which may be adaptive in the short-term (e.g., heightened attention to negative stimuli in a potentially dangerous situation), can undermine healthy psychological functioning over time. For instance, negativity biases in memory and attention are related to an increased risk for psychopathology, such as depression and anxiety [@mathews_cognitive_2005]. <!-- suggesting the importance of understanding the factors that contribute to individuals' biases. --> Alternatively, the malleability of valence bias offers insight into mechanisms or interventions that may be capable of shifting bias. Several brief, laboratory interventions are known to shift bias. For example, instructing participants to slow responding will lead to more positive interpretations of ambiguity [@neta_dont_2016]. Alternatively, manipulating participant gaze to match that of extremely negative (or positive) individuals modulates individuals' bias [@neta_through_2018]. In short, valence bias offers insight into both state- and trait-like components of individuals behavior, but through what mechanisms might an individual arrive at either a positive or negative interpretation?

## Initial negativity hypothesis
Myriad factors contribute to an individual's bias, but research suggests the initial interpretation is negative across people. Data supporting this initial negativity hypothesis come from many studies. As mentioned above, reaction times are faster for negative interpretations of ambiguous stimuli [@neta_dont_2016]. If negative interpretations occur faster than positive interpretations, then there may be an initial, automatic process resulting in the negativity. In turn, positive interpretations would require some override or regulation of this initial process. Supporting this idea, presentation of surprised facial expressions as low spatial frequency images, which are processed more readily than high spatial frequency images, bias interpretations towards negativity [@neta_primacy_2010]. Again, this supports the automaticity of negative interpretations, and suggests that high spatial frequency components, which are processed later, may be helpful for overriding negativity. <!-- Under this framework, arriving at a positive interpretation requires additional, top-down regulatory processes, and there is evidence to support this as well. For example, shifting participants time perspectives shifts interpretations of ambiguous images as well [@neta_its_2018].--> Evidence from the neuroimaging literature supports the initial negativity hypothesis as well; more positive individuals show higher levels of BOLD activation, when passively viewing surprised expressions, in brain regions recruited during emotion regulation of negative images [@petro_individual_2018]. Emotion regulation, and more specifically reappraisal, may be an underlying mechanism of a more positive valence bias.

<!-- Perceptual input also contributes to valence bias. In one recent study, Neta and colleagues [-@neta_all_2017] showed that faster intial fixation on the mouth is related to more positive interpretations of surprised faces.  In all, valence bias is a useful metric for understanding both trait-like components of individuals' affective biases, as well as gauging the effects of other experimental maniuplations on affective biases. -->
<!--For instance, a surprised expression could signal either a positive (e.g., winning the lottery) or negative (e.g., seeing a snake in the woods) affective state in the expresser. When contextual information is limited, individuals differ in their tendency to interpret surprised facial expressions as positive or negative. Importantly, this affective bias extends beyond facial expressions, as individuals often show a similar bias to surprised faces as they do for ambiguous scenes [@neta_neural_2013] or even words (Harp, Petro, Brown & Neta, in prep). This bias towards positive or negative interpretations is known as one's valence bias and myriad factors contribute to one's bias. -->
<!-- Both bottom-up (e.g., perceptual input) and top-down (e.g., stereotypes) processes are recruited for interpretation of facial expressions, and a growing body of work suggests that the initial interpretation of emotionally ambiguous stimuli is negative and driven by bottom-up processes. For instance, reaction times are faster for negative interpretations of ambiguous stimuli (just faces??) [@neta_dont_2016]. Additionally, presentation of surprised facial expressions as low spatial frequency images, which is processed more readily than high spatial frequency images, biased interpretations towards negativity [@neta_primacy_2010]. Consequently, under this framework, arriving at a positive interpretation requires additional, top-down regulatory processes and there is evidence to support this as well. For example, forcing participants to slow their responding during interpretations of ambiguous images shifts individuals' biases towards positivity [@neta_its_2018]. Neuroimaging evidence supports this initial negativity hypothesis as well, more positive individuals show higher levels of BOLD activation in brain regions recruited during emotion regulation [@petro_individual_2018]. <!-- Perceptual input also contributes to valence bias. In one recent study, Neta and colleagues [-@neta_all_2017] showed that faster intial fixation, as well as longer overall fixation, on the mouth is related to more positive interpretations of surprised faces and that forcing gaze patterns to match those of modulated interpretations of surprised expressions. THIS NEEDS TO BE REWORKED In short, slowing response time allows individuals to better regulate the potentially negative information in surprised expressions and to see it in a more positive light.  -->

## Cognitive loads and task interference
Executive functions, including working memory, are related to successful self-regulation, and in turn emotion regulation [@hofmann_executive_2012]. Directly comparing working memory and self-regulation of emotional responses, Schmeichel and colleagues [-@schmeichel_working_2008] reported that individuals with higher levels of working memory capacity demonstrated improved self-regulation towards the emotional stimuli. This suggests a connection--perhaps through some shared resource pool--between mitigated emotional responding and larger working memory resource availablility. Other work has focused on the effects of moods or affective states on working memory performance. For instance, some reports claim that both positive and negative mood interfere with working memory (Eyesenck and Calvo, 1992); however, others suggest benefits of positive mood on working memory [@yang_positive_2013]. Similarly, active working memory processes may alter concurrent affective processes. For instance, actively engaging working memory can mitigate emotional responses, particularly to negative stimuli [@macnamara_working_2011]. Recent neuroimaging work reports that negative emotional responses decrease as the cognitive demands of a working memory task increase [@van_dillen_tuning_2009]. Additionally, following an anger induction, those with low trait rumination show faster blood pressure recovery when provided with a distractor task [@gerin_role_2006]. Together, these studies suggest a resource competition between cognitive and emotional processes; in other words, when cognitive load demands are high (e.g., during active working memory maintenance), there are fewer resources available for other (i.e., affective) processes.

While previous work suggests an inhibitory effect on some emotional responses during periods with high cognitive demands, researchers have primarily focused on emotional responses to clearly valenced emotional stimuli. For instance, Schmeichel and colleagues [-@schmeichel_working_2008] showed participants videos intended to elicit strong negative responses (e.g., disgust) or positive responses (e.g., humor), while others have focused on comparing responses to neutral compared to negative sitmuli [@van_dillen_tuning_2009].<!--  Others have even measured symptoms during working memory maintenance of affective symptoms (i.e., typically increased negative affect), such as anxiety [@spangler_little_2017].--> However, many emotional appraisals, such as interpreting the valence of emotionally ambiguous stimuli, are more nuanced. <!-- For example, one may appraise the content of a billboard displaying a large order of french fries as either negative or positive depending on whether or not consuming that food is (in)congruent with one's current goals. This emotional appraisal is completed under concurrent load demands--that is, the perceiver must process both the emotional stimulus (i.e., the fries) as well as actively maintain their current goal state.--> Maintenance of goal states can reduce performance on some components of executive functioning tasks [@shaw_dieting_2004], which is in line with Lavie and colleague's [-@lavie_load_2004] load theory. Cognitive load theory posits that under a large cognitive load less executive resources are available to regulate incoming stimulus information. In turn, this means that more bottom-up, automatic processes (e.g., more readily processed perceptual input such as low spatial frequency information) might drive decisions under working memory loads.

## Domain-specific interference
Recently, cognitive load theory researchers have tested, and demonstrated, the domain-specificity of cognitive load and distractor interference effects in visual, spatial, and phonological domains [@burnham_components_2014]. Load interference effects may also transverse other domain components, such as maintaining emotional compared to neutral memory content. Emotional stimuli readily capture attention compared to neutral stimuli, and this is true even in participants with amygdala damage [@hodsoll_attentional_2011; @piech_attentional_2011]. Given the preference for emotional stimuli in the brain's information processing stream, it may be that cognitive loads with emotional content, compared to neutral, differentially affect concurrent emotional appraisals. Indeed, Kensigner and colleagues [-@kensinger_effect_2003] showed that negative emotional content slows performance on the n-back task. Other domain-specific effects have been observed in many lines of executive functions research, including those beyond the working memory domain. For example, the Stroop task [@stroop_studies_1935], a common measurement tool for inhibitory control, has been modified to include both emotional and non-emotional (neutral) stimuli [@whalen_emotional_2006] which has pronounced effects when the emotional words are population specific (e.g., trauma words in a PTSD sample). Other neuroimaging work also supports the notion that separate systems handle attentional biasing for domain-specific (emotional vs. non-emotional) task relevancy [@egner_dissociable_2008]. Given this evidence for dissociations of emotional and non-emotional information domains in executive functions (e.g., working memory, inhibitory control), the present work aims to clarify the interaction of emotional and non-emotional visual working memory demands on concurrent interpretations of emotional stimuli with ambiguous valence.

@@

## The present study
Given the evidence that positive interpretations of ambiguity rely on a regulatory mechanism, a demanding cognitive load could interfere with successful implementation of regulation strategies. Recently, Mattek and colleagues [-@mattek_differential_2016] showed that high levels of cognitive load (i.e., holding either a single or seven digit number in working memory) mitigates mouse trajectory deviations to modal responses during interpretations of surprised expressions. However, there was no effect on subjective valence interpretations. Further, <!-- Just as phonological and visual cognitive loads and distractors showed domain-specificity [@burnham_components_2014], and--> domain-specificity effects in both behavioral and BOLD data [@kensinger_effect_2003; @whalen_emotional_2006 ;@egner_dissociable_2008] suggest that a domain-specific (i.e., emotional) cognitive load may differentially affect emotional judgments during interpretations of ambiguity. <!--In other words, there is likely a relationship between the qualities of stimuli held in working memory and effects on concurrent task processing.--> Here, we aim to test the effects of low and high working memory loads in both emotional and neutral domains on valence bias. We expect that trials in which participants are maintaining an emotional working memory load will be more negative than neutral trials, as the high cognitive demand interferes with the regulatory mechanisms used to arrive at a positive interpretation. Further, we predict that higher working memory load trials, specifically in the emotional domain, will result in even more exaggerated negative interpretations.
